


\section{Ada Basics}

Ada is strongly typed with almost no type inference (e.g. let x = 10; compiler will not know that
x is an integer). Further it uses nominal typing. Therefore, even if 
two types have identical structure (e.g. are arrays of the same length, both are integer ranges), they
are considered distinct by Ada unless an explicit conversion is done by the programmer. For example

\begin{lstlisting}[caption={Conversion must be explicit}, label={con_expl}]
type Meters  is new Float;
type Seconds is new Float;

Distance : Meters  := 5.0;
Time     : Seconds := Seconds(Distance);  -- Explicitly convert Meters to Seconds
                                           -- Think of it like to_int(distance):
                                           -- a value must be converted to the target type
\end{lstlisting}

\noindent
In this example, \texttt{Meters} and \texttt{Seconds} are \emph{derived types}: that is, new types
created from an existing type (\texttt{Float}). Although they inherit the same operations and have the
same internal representation, they are considered distinct types and are not assignment-compatible.

\noindent
The variables \texttt{Distance} and \texttt{Time} are therefore of different types, even though they are both
represented using floating-point values. Ada prevents mixing such types implicitly. As a result, assigning
\texttt{Distance} to \texttt{Time} requires an explicit type conversion, as shown in the code. This does not
represent a meaningful physical conversion (meters to seconds) but simply demonstrates Ada's rule that
values of derived types must be explicitly converted when used in contexts expecting a different type.

\noindent Ada is a imperative and procedural language at its core (as opposed to declarative/functional).

\section{Build Alire project}

\noindent Inside appropriate folder run command "alr init chapter\_3", where chapter\_3 should be replaced with the name of the .gpr file one will use (if one wants a custom.gpr) and select library.

\noindent Move the toml to the main directory where the .gpr file is. Can delete all other created files by Alire. Then run command alr build.

